Retrofitting Effect Handlers onto OCaml

### 2020-11-11 retro-concurrency.tex
Updates

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
values, which when performed using the |perform| primitive returns a |'a|
value. Constructors for the value of type |'a eff| are declared using the
effect declarations. For example, the declaration |effect E : string -> int|
is syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be |continue|d with a
suitably typed value or |discontinue|d with an exception.

@@

We present the operational semantics for a core language of effect handlers in
Multicore OCaml. The operational semantics intends to faithfully capture the
semantics of Multicore OCaml effect handlers, and to inform an implementation.

@@

Figure~\ref{sem:syntax} presents the syntax of expressions and configurations.
Our expressions consists of integer constants ($n$), variables ($x$),
abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic expressions ($e
\odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising exceptions
($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and handling effects
($\handle{e}{h}$). Abstractions come in two forms: OCaml abstractions
($\lambda^o$) and C abstractions ($\lambda^c$). The handler consists of a
return case ($\caseval{x}{e}$), zero or more exception cases
($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
continuation $k$ and body $e$.

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen86} for effect handlers, following the abstract
machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
difference from Hillerstrom et al. is that our stacks are composed of
alternating sequence of OCaml and C stack segments. The program state is
captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
the current term $\tau$ under evaluation, its environment $\env$ and the
current stack $\stack$. The term is either an expression $e$ or a value $v$.
The value are integer constants $n$, continuations $k$, closure
$\clos{x}{e}{\env}$, effect being performed $\effval{l}{k}$ and exception being
raised $\exnval{l}$. The environment is a map from variables to values.

The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the
C stack $\cstack$ under it. Thus, the program stack is an alternating sequence
of C and OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The
frame list $\fl$ is composed of individual frames $r$, which is one of an
argument frame $\farg{e}{\env}$ with the expression $e$ at the argument
position of an application with its environment $\env$, a function frame
$\ffun{v}$ with the value at the function position of an application, and
frames for evaluating the arguments of an arithmetic expression.

@@

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel}) can be
performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
Intuitively, this splits the world into executing C code and OCaml code.

@@

The C reduction steps are given in Figure~\ref{sem:cstep}. We can take a C step
by taking an administrative reduction step $\rightsquigarrow$. The
administrative reductions are common to both C and OCaml. The rules
\textsc{Var}, \textsc{Arith1}, \textsc{Arith2}, \textsc{App1}, \textsc{App2}
and \textsc{App3} are standard. \textsc{Arith3} performs the arithmetic
operation on the integers ($\llbracket n_1 \odot n_2 \rrbracket$).
\textsc{Raise} pushes an function frame with exception value to indicate that
an exception is being raised. Similarly, \textsc{Perform} pushes a function
frame with an effect value with an empty continuation
$[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
handler with an identity return case alone. We shall return to \textsc{Resume1}
and \textsc{Resume2} later.

@@

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fibers only involves saving the exception and the
stack pointer of the current stack and loading the same on the target stack.
Since OCaml does not generate pointers into the stack, the two |fiber_info|
fields are the only ones that need to be updated when fibers are moved. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we observed
that directly using |mimalloc|~\cite{Leijen19} outperforms having the stack
cache.

@@

Recall that OCaml programs are written with the expectation that function calls
return exactly once (\S\ref{sec:refine}). Consider the scenario when a
continuation is neither |continue|d nor |discontinue|d. Since fibers are
|malloc|ed and are |free|d when the computation being handled returns a value
or raises an exception, not resuming continuations leaks memory. In addition,
unresumed continuations may also leak other resource such as open file
descriptors.

@@

Multicore OCaml effect handlers are similar to~\cite{Leijen14, Kiselyov12}, but
are more efficient; since we only allow our continuations to be resumed at-most
once, there is no need for copying frames. Bruggeman et al.~\cite{Bruggeman96}
show how to implement one-shot continuations efficiently using segmented stacks
in Scheme. Farvardin et al.~\cite{Farvardin20} perform a comprehensive
evaluation of various implementation strategies for continuations on modern
hardware. Multicore OCaml stacks do not neatly fit the description of one of
these implementation strategies. They best described as using the resize
strategy from Farvardin et al. for each of the fibers, which are in turn linked
to represent the current stack and the captured continuations.

@@

The interaction of non-local control flow and resources has been studied
extensively. Scheme uses |dynamic-wind|~\cite{}, which is a generalisation of
Common Lisp |unwind-protect|~\cite{}, which ensures de-allocation (and
re-allocation) of resources every time the non-local control leaves and enters
back into a context. |dynamic-wind| is considered to be complex and not quite
what one wants for resource clean up~\cite{Kiselyov,Sitaram03}. The main
challenge is distinguishing returning exits from non-returning ones. Resource
clean up needs to occur only during non-returning exit.

@@

Leijen~\cite{Leijen18} explicitly extends effect handlers with |initially| and
|finally| clauses in Koka for resource safety. Sivaramakrishnan et
al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
exceptions. This is orthogonal to the contributions of this paper. Our focus i
the compiler and runtime system support for implementing effect handlers.

Languages with effect handlers typically allow continuations to be resumed more
than once to enable use cases such as backtracking search and probabilistic
programming~\cite{Pyro}. Multicore OCaml has an experimental feature for
explicitly cloning continuations, which makes a copy of each of its constituent
fibers. Multi-shot continuations are difficult to reason about in the presence
of linear resources such as file descriptors. Recall that OCaml programs are
written with the expectation that function calls return \emph{exactly-once}
either normally or exceptionally. With multi-shot continuations, they may
return \emph{more than once}, leading to errors such as double-freeing
malloc'ed memory. Since our goal is to retrofit effect handlers onto OCaml, we
relegate continuation cloning to the OCaml |Obj| module, which contains unsafe
features.

>>>>

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{ExtVariants} |'a eff| of effect
values, which when performed using the |perform| primitive returns a |'a|
value. Constructors for the value of type |'a eff| are declared using the
effect declarations. For example, the declaration |effect E : string -> int|
is syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be continued with a
suitably typed value or discontinued with an exception.

@@

We present the operational semantics for a core language of effect handlers.
The operational semantics intends to faithfully capture the semantics of
Multicore OCaml effect handlers, and to inform an implementation.

@@

Our expressions (Figure~\ref{sem:syntax}) consists of integer constants ($n$),
variables ($x$), abstraction ($\lam{x}{x}$), application ($e~e$), arithmetic
expressions ($e \odot e$) where $\odot$ ranges over $\{+,-,*,/\}$, raising
exceptions ($\throw{l}{e}$), performing effects ($\perform{l}{e}$), and
handling effects ($\handle{e}{h}$). Abstractions come in two forms: OCaml
abstractions ($\lambda^o$) and C abstractions ($\lambda^c$). The handler
consists of a return case ($\caseval{x}{e}$), zero or more exception cases
($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and body $e$, and zero or
more effect cases ($\caseeff{l}{x}{k}{e}$) with label $l$, parameter $x$,
continuation $k$ and body $e$.

The operational semantics is an extension of a CEK machine
semantics~\cite{Felleisen86} for effect handlers, following the abstract
machine semantics of Hillerstrom et al.~\cite{Hillerstrom20}. The key
difference from Hillerstrom et al. is that our stacks are composed of
alternating sequence of OCaml and C stack segments. The program state is
captured as configuration $\config \coloneqq \configa{\tau}{\env}{\stack}$ with
the current term $\tau$ under evaluation, its environment $\env$ and the
current stack $\stack$. The term is either an expression $e$ or a value $v$.
The value are integer constants $n$, continuations $k$, closure
$\clos{x}{e}{\env}$, effect being performed ($\effval{l}{k}$) and exception
being raised ($\exnval{l}$). The environment is a map from variables to values.

The stack $\stack$ is either a C stack ($\cstack$) or an OCaml stack
($\ostack$). Every C stack $\cstacka{\fl}{\ostack}$ consists of a list of
frames $\fl$, and the OCaml stack $\ostack$ under it. Every OCaml stack is
either empty $\ostackemp$ or non-empty $\ostacka{k}{\cstack}$ with the C stack
$\cstack$ under it. Thus, the program stack is an alternating sequence of C and
OCaml stacks terminating with an empty OCaml stack $\ostackemp$. The frame list
$\fl$ is composed of individual frames $r$, which is one of an argument frame
$\farg{e}{\env}$ with the expression $e$ at the argument position of an
application with its environment $\env$, a function frame $\ffun{v}$ with the
value $v$ at the function position of an application, and frames for evaluating
the arguments of an arithmetic expression.

@@

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions (Figure ~\ref{sem:toplevel}) can be
performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.

@@

We can take a C step (Figure~\ref{sem:cstep}) by taking an administrative
reduction step $\rightsquigarrow$. The administrative reductions are common to
both C and OCaml. The rules \textsc{Var}, \textsc{Arith1}, \textsc{Arith2},
\textsc{App1}, \textsc{App2} and \textsc{App3} are standard. \textsc{Arith3}
performs the arithmetic operation on the integers ($\llbracket n_1 \odot n_2
\rrbracket$). \textsc{Raise} pushes an function frame with exception value to
indicate that an exception is being raised. Similarly, \textsc{Perform} pushes
a function frame with an effect value with an empty continuation
$[[],(\{\caseval{x}{x}\},\emptyset)]$ with no captured frames and an empty
handler with an identity return case alone. We shall return to \textsc{Resume1}
and \textsc{Resume2} later.

@@

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fibers only involves saving the exception and the
stack pointer of the current stack and loading the same on the target stack.
Since OCaml does not generate pointers into the stack, the two |fiber_info|
fields are the only ones that need to be updated when fibers are moved. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation~\cite{Farvardin20}. In our experiments, we
observed that directly using |mimalloc|~\cite{Leijen19} outperforms the stack
cache.

@@

Recall that OCaml programs are written with the expectation that function calls
return exactly once (\S\ref{sec:refine}). Consider the scenario when a
continuation is neither continued nor discontinued. Since fibers are
malloced and are |free|d when the computation being handled returns a value
or raises an exception, not resuming continuations leaks memory. In addition,
unresumed continuations may also leak other resource such as open file
descriptors.

@@

Similar to~\cite{Leijen14, Kiselyov12}, Multicore OCaml uses the call stack for
implementing continuations, but our continuations are one-shot. Bruggeman et
al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
comprehensive evaluation of various implementation strategies for continuations
on modern hardware. Multicore OCaml stacks do not neatly fit the description of
one of these implementation strategies. They best described as using the resize
strategy from Farvardin et al. for each of the fibers, which are in turn linked
to represent the current stack and the captured continuations.

@@

Multicore OCaml is not the first language to support stack inspection in the
presence of non-local control operators. Chez Scheme supports continuation
marks~\cite{Flatt20} which permit stack inspection to implement dynamic
binding, exceptions, profilers, debuggers, etc, in the presence of first-class
continuations. As the authors note, continuation marks can be implemented using
effect handlers, but direct support for continuation marks leads to better
performance. In this work, we focus on retaining the support for stack
inspection through DWARF unwind tables in the presence of effect handlers.

The interaction of non-local control flow and resources has been studied
extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation of
Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation (and
re-allocation) of resources every time the non-local control leaves and enters
back into a context. |dynamic-wind| is not quite what one wants for resource
clean up~\cite{Kiselyov,Sitaram03}. The main challenge is distinguishing
returning exits from non-returning ones. Resource clean up needs to occur only
during non-returning exit.

@@

Leijen~\cite{Leijen18} explicitly extends effect handlers with |initially| and
|finally| clauses in Koka for resource safety. Sivaramakrishnan et
al.~\cite{TFP17} describe the interaction of effect handlers and asynchronous
exceptions. This is orthogonal to the contributions of this paper. Our focus is
the compiler and runtime system support for implementing effect handlers.

Languages with effect handlers typically allow continuations to be resumed more
than once to enable use cases such as backtracking search and probabilistic
programming~\cite{Pyro}. Multicore OCaml has an experimental feature for
explicitly cloning continuations, which makes a copy of each of its constituent
fibers. Multi-shot continuations are difficult to reason about in the presence
of linear resources such as file descriptors. Recall that OCaml programs are
written with the expectation that function calls return \emph{exactly-once}
either normally or exceptionally. With multi-shot continuations, they may
return \emph{more than once}, leading to errors such as double-freeing
malloced memory. Since our goal is to retrofit effect handlers onto OCaml, we
relegate continuation cloning to the OCaml |Obj| module, which contains unsafe
features.
